---
title: "NFL Data Bowl Practice"
author: "Dominic Hoar-Weiler"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Click to clear environment
rm(list = ls())
```

#Importing Data

```{r}
# Increase memory size (Windows only)
memory.limit(size = 64000)  # Adjust as needed

library(tidyverse)
player_raw_data <- read_csv("players.csv")
plays_raw_data <- read_csv("plays.csv")
games_raw_data <- read_csv("games.csv")
# List all files matching the pattern "week" (e.g., "tracking_week_1.csv")
files <- list.files(pattern = "week", full.names = TRUE)
tracking_list <- list()

```
```{r}
for (file in files) {
  tracking <- read_csv(file)
  tracking <- tracking |>
    mutate(
    # make all plays go from left to right
    x = ifelse(playDirection == "left", 120 - x, x),
    y = ifelse(playDirection == "left", 160 / 3 - y, y),
    # flip player direction and orientation
    dir = ifelse(playDirection == "left", dir + 180, dir),
    dir = ifelse(dir > 360, dir - 360, dir),
    o = ifelse(playDirection == "left", o + 180, o),
    o = ifelse(o > 360, o - 360, o)
    )
  tracking_list[[file]] <- tracking
}
```

# Play Clock at Snap Exploration
  •	Topic: Predict play outcomes by the moment at which the ball is snapped
  each play
	•	Key Questions:
	  * What pre-snap factors have an influence on when the ball is snapped?
	  * How does the play clock at snap relate to play success?
	•	Deliverable: Conclusion about the importance of play clock at snap and how
	  teams should approach this piece of data
	  
```{r}
plays_raw_data$passResult_enc <- as.numeric(factor(plays_raw_data$passResult)) - 1
  
playclock.lm <- glm(passResult_enc ~ playClockAtSnap, family = binomial(link = "logit"),
                    data = plays_raw_data, na.action = na.omit)
summary(playclock.lm)

```


This is the most interesting find in my data. I categorized the playClockAtSnap 
into high, medium, and low. I then plotted the results of the different buckets 
to see the results. We can see that the lower the play clock, might have a 
higher likelihood of rushed plays and false starts. I was thinking that it could 
be correlated with poor play outcomes due to rushed decision making 
or how ready the defense is. On the other hand, I theorized that the higher play 
clock could be correlated with ready offenses while also potential catching
the defense off guard. We can see that when the playClock is less than or equal to 
5s, the results of the play are more spread out than than when the playClock is higher
indicating that results vary more.
```{r}
# Categorize playClockAtSnap into "High," "Medium," and "Low"
plays_raw_data <- plays_raw_data %>%
  mutate(playClockCategory = case_when(playClockAtSnap >= 15 ~ "playClock More Than 15s",               
      playClockAtSnap >= 5 & playClockAtSnap < 15 ~ "playClock Between 5s and 15s", 
      playClockAtSnap < 5 ~ "playClock Less Than 5s",               
    )
  )

plays_filtered <- plays_raw_data %>%
  filter(!is.na(playClockCategory))

ggplot(plays_filtered, aes(x = playClockCategory, fill = passResult)) +
  geom_bar(position = "fill") +
  labs(
    title = "Play Outcomes by Play Clock Category",
    x = "Play Clock Category",
    y = "Proportion",
    fill = "Pass Result"
  ) +
  theme_minimal()
```
And we can see here, the playClock is indeed statistically significant with a
value of 0.0002549 < 0.05. The next step would be to find a model to potentially 
predict results.
```{r}
contingency_table <- table(plays_filtered$playClockCategory, plays_filtered$passResult)

chi_sq_test <- chisq.test(contingency_table)
print(chi_sq_test)
```

# Pre-Snap Blitz Detection
	•	Topic: Predict defensive blitzes based on pre-snap player positioning and 
	movement.
	•	Key Questions:
	•	What pre-snap defensive alignments or movements are the best indicators of 
	a blitz?
	•	How does successful blitz detection affect QB performance and play outcomes?
	•	Deliverable: A machine learning model to predict blitz likelihood and its 
	relationship with play success.
	
One question we have to answer is how do we identify blitzes? The common 
definition of a blitz is when the defense sends more than the usual amount of 
defenders to rush the quarterback. The three types of blitzes are corner 
blitzes, safety blitzes and linebacker blitzes. These types simply identify 
what positions are sent to blitz.

Below, I'm attempting to trim the tracking data, and then add columns so that
we can measure whether players are rushing or not, so that we can then set
a threshold of number of rushers, to identify if plays are blitzes or not.
	
```{r}
dom_mov_data <- tracking_week_1 |>
  group_by(gameId, playId, nflId) |>
  left_join(player_raw_data, by = c("nflId"))

tracking_qb <- dom_mov_data |>
  filter(position == "QB") |>
  select(gameId, playId, frameId, x_qb = x, y_qb = y)

movement_data <- dom_mov_data |>
  filter(frameType == "BEFORE_SNAP") |>
  group_by(gameId, playId, nflId) |>
  summarize(
    total_movement = sum(sqrt((x - lag(x, default = first(x)))^2 + (y - lag(y, default = first(y)))^2), na.rm = TRUE),
    .groups = "drop"  # Explicitly drop grouping
  )

tracking_pass_rush <- dom_mov_data |>
  filter(position %in% c("SS","FS","CB","ILB","OLB")) |>
  left_join(tracking_qb, by = c("gameId","playId","frameId")) |>
  group_by(gameId, playId, frameId) |>
  mutate(d_qb = sqrt((x - x_qb) ^ 2 + (y - y_qb) ^ 2),
         v_qb = -(d_qb - lag(d_qb)) / 0.1) |>
  ungroup() |>
  left_join(movement_data, by = c("gameId","playId","nflId.x" = "nflId"))

tracking_pass_rush <- tracking_pass_rush |>
  select(gameId,playId,frameId,nflId.x,displayName.x,frameType,time,playDirection,
         x,y,s,a,dis,o,dir,event,position,x_qb,y_qb,d_qb,v_qb,total_movement) |>
   mutate(blitzing = ifelse((dir > 180 | d_qb < 8) & v_qb > 5 & frameType == "AFTER_SNAP",TRUE,FALSE))

# Remove rows with NA values in blitzing
tracking_pass_rush_clean <- tracking_pass_rush |>
  left_join(plays_raw_data |> select(gameId, playId, offenseFormation, playClockAtSnap, pff_passCoverage,down,yardsToGo), by = c("gameId", "playId")) |>
  mutate(motion_flag = ifelse(total_movement > 3, 1, 0)) |>
  filter(
    !is.na(blitzing),
    !is.na(total_movement),
    !is.na(offenseFormation),
    !is.na(playClockAtSnap)
  )
```

Now that we've filtered the data accordingly, let's create the models.

```{r}
# Check updated data
head(tracking_pass_rush_clean)

# Create model
predicting_blitz.blogr <- glm(blitzing ~ total_movement + offenseFormation + playClockAtSnap, 
                              data = tracking_pass_rush_clean, 
                              family = binomial(link = "logit"))

predicted_probs <- predict(predicting_blitz.blogr, type = "response")

predicting_blitz <- glm(blitzing ~ total_movement + 
                                  offenseFormation + 
                                  down + yardsToGo + 
                                  motion_flag + 
                                  d_qb,
                        family = binomial(link = "logit"),
                        data = tracking_pass_rush_clean)

# Test if the data are logistic
library(ResourceSelection)
hoslem.test(tracking_pass_rush_clean$blitzing,predicted_probs)
```

Since the Hosmer and Lemeshow goodness of fit test yielded a p-value of
$2.2 * 10^{-6}$. This means that we reject the null hypothesis that the data are 
reasonably logistic. This means that the logistic model is likely not a good
fit for the data.

That being said, let's proceed anyway to see what we can find.

```{r}
# Summarize both models
summary(predicting_blitz.blogr)
summary(predicting_blitz)
# Found correlation between pre-snap movement and whether or not that player blitzed
```

Based on the above information, we have that it is possible that there is a 
relationship between total pre-snap movement of a player and whether they blitz
or not. Also, by running the Hosmer and Lemeshow test, we found that the data 
are not reasonably logistic, meaning the resulting data from the binary 
logistic regression model is unreliable.

Here we try other simple machine learning classification models such as Linear 
and Quadratic Discriminant Analysis (LDA & QDA) and classification trees with the
same predictor variables as before.

#LDA and QDA:
```{r}
library(MASS)

# Linear Discriminant Analysis
blitz_lda <- lda(blitzing ~ total_movement + offenseFormation + down + yardsToGo,
                 data = tracking_pass_rush_clean)
print(blitz_lda)

# Quadratic Discriminant Analysis
blitz_qda <- qda(blitzing ~ total_movement + offenseFormation + down + yardsToGo,
                 data = tracking_pass_rush_clean)
print(blitz_qda)

# Predictions
lda_pred <- predict(blitz_lda)$class
qda_pred <- predict(blitz_qda)$class

# Confusion matrix
table(Predicted = lda_pred, Actual = tracking_pass_rush_clean$blitzing)
table(Predicted = qda_pred, Actual = tracking_pass_rush_clean$blitzing)
```

From the two discriminant analysis models above we can see that the players who 
blitz (TRUE) have lower total movement (6.40 vs. 8.65 for FALSE). However, similar
to the logistic regression model, it identifies some patterns but the coefficients
and group mean indicate that these differences are subtle.

#Classification Tree:
```{r}
library(rpart)
library(rpart.plot)

# Fit a classification tree
blitz_tree <- rpart(blitzing ~ total_movement + offenseFormation + down + yardsToGo, 
                    data = tracking_pass_rush_clean, method = "class")

# Plot the tree
rpart.plot(blitz_tree, main = "Classification Tree for Blitz Prediction", type = 2)

# Predictions
tree_pred <- predict(blitz_tree, type = "class")

# Confusion matrix
table(Predicted = tree_pred, Actual = tracking_pass_rush_clean$blitzing)

```

From the tree graphic, we can see a single node predicting FALSE which means that
the model predicts all observations as FALSE (no blitzes detected). This makes sense
because the probability of the blitzing is only 0.09% in the actual data indicating
that classification tree would not be the best at splitting the data well enough
to predict. 
